---
title: "Inpatient Capacity in the U.S. During the 2020 COVID-19 Crisis"
author: "Brian Murphy"
date: "8/13/2020"
output: slidy_presentation
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
```

## Initialize Libraries
```{r}
library(RCurl)
library(tidyverse)
library(data.table)
library(ComplexHeatmap)
library(scales)
library(directlabels)
library(grid)
library(openintro)
library(car)
```

## Background: Data

COVID-19 Estimated Patient Impact and Hospital Capacity by State

### Source:

* U.S. Department of Health and Human Services (HHS) 
* HHS TeleTracking 
* Reporting provided directly to HHS Protect by state/territorial health departments on behalf of their healthcare facilities
* https://healthdata.gov/dataset/covid-19-estimated-patient-impact-and-hospital-capacity-state

### Estimation Method:

* Three metrics of interest
  + Estimated inpatient beds occupied
  + Estimated inpatient beds occupied by COVID-19 patients
  + Estimated ICU beds occupied
* Data are daily, by state in the U.S. in 2020
* Reported occupied beds
  + Percentage of beds occupied for reporting hospitals
  + Occupied beds / Total beds
* Total inpatient beds
  + Calculate weights based on percentage of total beds associated with reporting hospitals
  + Estimate total inpatient beds by dividing total beds by reporting hospital weight
* Expected unreported occupied beds = Estimated total inpatient beds - Reported occupied beds
* Total estimated occupied beds = Reported occupied beds + Expected unreported occupied beds

### Import Inpatient Beds Data
```{r}
# Inpatient Occupied
url <- "https://raw.githubusercontent.com/bmurphy1993/NYU-R-Programming-Final-Project/master/reported_inpatient_all_20200720_0537.csv"
inpatient <- read.csv(url, header = TRUE, sep = ",", quote = "\"'")

# COVID-19 Inpatient Occupied
url <- "https://raw.githubusercontent.com/bmurphy1993/NYU-R-Programming-Final-Project/master/inpatient_covid_final_20200720_0537.csv"
covid <- read.csv(url, header = TRUE, sep = ",", quote = "\"'")

# ICU Occupied
url <- "https://raw.githubusercontent.com/bmurphy1993/NYU-R-Programming-Final-Project/master/icu_final_20200720_0537.csv"
icu <- read.csv(url, header = TRUE, sep = ",", quote = "\"'")
```

***

## Background: Data

### Variables

Estimated inpatient beds occupied
```{r}
summary(inpatient)
```

Estimated inpatient beds occupied by COVID-19 patients
```{r}
summary(covid)
```

Estimated ICU beds occupied
```{r}
summary(icu)
```

***

## Background: Analysis

### Analysis Questions/Potential Insights

* Have U.S. hospitals been pushed beyond normal inpatient and ICU capacity during the COVID-19 crisis?
  + Hypothesis: Yes for both metrics
* Whether and to what extent U.S. hospitals have expanded their normal "capacity"
* Have COVID-19 inpatients contributed significantly to these changes?
* Do the changes appear to be temporary or will they have lasting effects on hospital capacity?

***

## Methods: Data Cleaning

Rename common variables
```{r}
# Rename common variables so that they will remain distinct after merging data
df_list = list(inpatient = inpatient, covid = covid, icu = icu)
for(df_name in names(df_list)) {
  df_list[[df_name]] = df_list[[df_name]] %>%
    rename_with(.fn = ~paste(., df_name, sep = "."), .cols = c("Count.LL", "Count.UL", "Percentage.LL", "Percentage.UL"))
}
```
Merge the data sets in the list
```{r}
# Merge by "state" and "collection_date"
beds_all <- df_list %>% 
  reduce(left_join, by = c("state", "collection_date"))
```
Reformat the date variable
```{r}
# Reformat "collection_date" variable from character to date
beds_all$collection_date <- as.Date(beds_all$collection_date, "%Y-%m-%d")
```
Final cleanup
```{r}
# Rename long variable names
setnames(beds_all, 
         old = c("Inpatient.Beds.Occupied.Estimated","Percentage.of.Inpatient.Beds.Occupied.Estimated", "Inpatient.Beds.Occupied.by.COVID.19.Patients.Estimated", "Percentage.of.Inpatient.Beds.Occupied.by.COVID.19.Patients.Estimated", "ICU.Beds.Occupied.Estimated", "Percentage.of.ICU.Beds.Occupied.Estimated"), 
         new = c("inpatient.occupied","inpatient.percent","covid.occupied", "covid.percent","icu.occupied", "icu.percent"))

# Drop the upper and lower limit variables
beds <- beds_all[,-c(4:5, 7:8, 10:11, 13:14, 16:17, 19:20)]

head(beds)
```
***

## Methods: Data Exploration
Summary info
```{r}
summary(beds)
  # Max inpatient and covid occupancy = 100%
  # Max ICU capacity = Inf. Replace Inf with NA.
beds$icu.percent <- na_if(beds$icu.percent, Inf)
summary(beds$icu.percent)
```
***

## Methods: Data Exploration

Deal with missing values
```{r}
beds[is.na(beds[6]),]
  # Captures missing values for inpatient, inpatient.percent, covid, and covid.percent
  # These are only HI and GU. Only interested in mainland U.S. because it is unlikely that 
unique(beds$state)
  # Drop non-mainland states and territories: Alaska (AK), Hawaii (HI), Curacao (CW), Guam (GU), Puerto Rico (PR) 
beds_main <- beds[ ! beds$state %in% c("AK", "HI", "CW", "GU", "PR"), ]
unique(beds_main$state) # Now those states and territories are dropped

head(beds_main[is.na(beds_main[7]),])
beds_main[is.na(beds_main$icu.occupied),] %>% 
  group_by(state) %>% 
  summarise(max_date = max(collection_date)) 
  # For the most part, looks like data  was not tracking ICU numbers earlier in the pandemic for certain states. Leave NAs alone for now.
```
***

## Methods: Data Transformations

### New Variables

Calculate max capacity for each metric
```{r}
# Use 'transform()' function to calculate capacity = occupied / percent_occupied
beds_main <- transform(beds_main, inpatient.capacity = inpatient.occupied / inpatient.percent)
beds_main <- transform(beds_main, covid.capacity = covid.occupied / covid.percent)
beds_main <- transform(beds_main, icu.capacity = icu.occupied / icu.percent)
  # COVID capacity should equal inpatient capacity. Sometimes does, sometimes doesn't. Coming back to this.

# The new capacity variables sometimes equal "NaN" because of denominators equal to zero
  # Change NaNs to missing using 'na_if()' function
beds_main$inpatient.capacity <- na_if(beds_main$inpatient.capacity, "NaN")
beds_main$covid.capacity <- na_if(beds_main$covid.capacity, "NaN")
beds_main$icu.capacity <- na_if(beds_main$icu.capacity, "NaN")
```

Investigate inpatient capacity further
```{r}
# Compare inpatient capacity and COVID-19 capacity. These should be the same in theory.
beds_main %>% 
  group_by(state) %>% 
  summarise(mean_inpat = mean(inpatient.capacity, na.rm = TRUE), mean_cov = mean(covid.capacity, na.rm = TRUE),
            med_inpat = median(inpatient.capacity, na.rm = TRUE), med_cov = median(covid.capacity, na.rm = TRUE))
```
Drop covid.capacity and use inpatient capacity
```{r}
beds_main <- beds_main[,-10]
```
* In theory, they estimate the same thing
* covid.capacity has a fair amount of missing values
* For non-missing values, means and medians are similar
* Differences are likely due to the fact that all these metrics are estimated



Can recalculate COVID % of capacity based on inpatient capacity:
```{r}
beds_main$covid.percent <- beds_main$covid.occupied / beds_main$inpatient.capacity
  # This allows for apples-to-apples comparisons of COVID-19 and all inpatient occupancy %s 
```

***

## Methods: Data Transformations

### Create Transformed Data Cuts for Visualizations

Collapse to Entire U.S.
```{r}
# Use this to see nationwide change over time
beds_main_total <- beds_main %>% 
  group_by(collection_date) %>% # Summing by date essentially removes the state variable to get daily totals for the entire country
  summarise(inpatient.occupied = sum(inpatient.occupied, na.rm = TRUE), 
            covid.occupied = sum(covid.occupied, na.rm = TRUE), 
            icu.occupied = sum(icu.occupied, na.rm = TRUE), 
            inpatient.capacity = sum(inpatient.capacity, na.rm = TRUE), 
            icu.capacity = sum(icu.capacity, na.rm = TRUE))
```
***

## Methods: Data Transformations

### Create Transformed Data Cuts for Visualizations

Matrices for heatmaps: States as columns, days as rows
```{r}
# Inpatient Percent of Capacity
inpat.percent <- spread(beds_main[,c(1:2,4)],state,inpatient.percent) # spread() function reshapes data
inpat.percent.mat <- as.matrix(inpat.percent[,2:50]) # turn into matrix for heatmap
rnames <- as.character(inpat.percent[,1]) # Use dates as row names in matrix
rownames(inpat.percent.mat) <- rnames

tail(inpat.percent.mat) # Used 'tail()' rather than 'head()' because many early values are missing

# COVID Percent of Capacity
cov.percent <- spread(beds_main[,c(1:2,6)],state,covid.percent)
cov.percent.mat <- as.matrix(cov.percent[,2:50])
rnames <- as.character(cov.percent[,1])
rownames(cov.percent.mat) <- rnames

# ICU Percent of Capacity
ICU.percent <- spread(beds_main[,c(1:2,8)],state,icu.percent)
ICU.percent.mat <- as.matrix(ICU.percent[,2:50])
rnames <- as.character(ICU.percent[,1])
rownames(ICU.percent.mat) <- rnames
```

***

## Results: Visualizations

### Plot the variables in question for the entire U.S.
```{r}
# Add one more variable to show the total inpatients not including COVID-19
beds_main_total <- transform(beds_main_total, inpatient.non_covid = inpatient.occupied - covid.occupied)
# Reshape data using 'melt()' function to use for a ggplot2 graph
beds_main_total_melt <- reshape2::melt(beds_main_total, id.var='collection_date')

# Create line plot to show development over the course of the year
lp <- ggplot(beds_main_total_melt, aes(x = collection_date, y = value, col = variable)) + # Uses "melted" data to plot all variables at once
  geom_line(size = 1, aes(linetype = variable)) + 
  scale_linetype_manual(values=c("solid", "solid", "solid", "dashed", "dashed", "solid")) + # Make capacities dashed
  labs(title = "U.S. Hospital Bed Occupancy in 2020", x = "Date in 2020", y = "Total Beds") +
  theme(plot.title = element_text(hjust = 0.5), legend.position = "none", plot.margin = unit(c(1,6,1,1), "lines")) +
  # Add axis and series labels
  scale_y_continuous(labels = number) +
  geom_text(data = subset(beds_main_total_melt, collection_date == 18457), aes(label = variable), hjust = -.1) +
  scale_colour_discrete(guide = 'none') +
  scale_x_date(date_labels = "%b", date_breaks = "1 month")

# Allow series labels to hang off the chart
gt <- ggplotGrob(lp)
gt$layout$clip[gt$layout$name == "panel"] <- "off"
grid.draw(gt)
```

***

## Results: Visualizations

### Heatmaps: Hospital Bed Occupancy by State

Could use line plots, but they get crowded with 50 series. Heatmaps help with this and can still show change over time
```{r}
# Inpatient Percent of Capacity
Heatmap(inpat.percent.mat, 
        row_title = "Date Ascdending from Top", 
        column_title = "Inpatient Percent of Total Beds by State in 2020",
        cluster_rows = FALSE, # Want rows to remain sorted by date
        show_row_names = FALSE,
        show_column_dend = FALSE,
        column_names_gp = gpar(fontsize = 10),
        heatmap_legend_param = list(title = "")
        )

# COVID Percent of Capacity
Heatmap(cov.percent.mat, 
        row_title = "Date Ascdending from Top", 
        column_title = "COVID-19 Percent of Total Beds by State in 2020",
        cluster_rows = FALSE,
        show_row_names = FALSE,
        show_column_dend = FALSE,
        column_names_gp = gpar(fontsize = 10),
        heatmap_legend_param = list(title = "")
        )

# ICU Percent of Capacity
Heatmap(ICU.percent.mat, 
        row_title = "Date Ascdending from Top", 
        column_title = "Percent of Total ICU Beds by State in 2020",
        cluster_rows = FALSE,
        show_row_names = FALSE,
        show_column_dend = FALSE,
        column_names_gp = gpar(fontsize = 10),
        heatmap_legend_param = list(title = "")
        )

# Column clustering allows us to approximate which states have been pushed closest to capacity
# Maps also show the initial "wave" of cases in some states. Reflects that states were initially pushed to their limits but then expanded capacity
# Interestingly, the states with the highest overall occupancy % did not necessarily have highest ICU occupancy
```

***

## Results: Visualizations

### Heatmaps with pre-COVID-19 Hospital Capacities

Redefine Inpatient capacity and ICU capacity to their initial values
```{r}
# Inpatient original cap
  # Using 'which()' and 'ave()' functions, find the inpatient capacity for each state at the earliest date available 
  # in the data to estimate what capacity would have been before the invlux of COVID-19 cases  
beds_main_inpat_nona <- beds_main[!is.na(beds_main$inpatient.capacity),]
original_cap <- beds_main_inpat_nona[
  which(beds_main_inpat_nona$collection_date == 
          ave(beds_main_inpat_nona$collection_date, beds_main_inpat_nona$state, FUN = min)), c(1, 9)]

# ICU original cap
beds_main_icu_nona <- beds_main[!is.na(beds_main$icu.capacity),]
original_cap_icu <- beds_main_icu_nona[
  which(beds_main_icu_nona$collection_date == 
          ave(beds_main_icu_nona$collection_date, beds_main_icu_nona$state, FUN = min)), c(1, 10)] 

# Merge the inpatient and ICU initial capacities with 'left_join()' funciton
original_cap <- left_join(original_cap, original_cap_icu, by = "state")

original_cap <- rename(original_cap, "inpat.cap.orig" = inpatient.capacity)
original_cap <- rename(original_cap, "icu.cap.orig" = icu.capacity)

# Merge back onto main dataset 
beds_main_origcap <- left_join(beds_main, original_cap, by = "state") 
beds_main_origcap$icu.cap.orig <- ifelse(beds_main_origcap$icu.occupied == 0, NA, beds_main_origcap$icu.cap.orig)

# Redefine %s as portion of initial capacities rather than the dynamic capacities
beds_main_origcap$inpatient.percent <- beds_main_origcap$inpatient.occupied / beds_main_origcap$inpat.cap.orig
beds_main_origcap$covid.percent <- beds_main_origcap$covid.occupied / beds_main_origcap$inpat.cap.orig
beds_main_origcap$icu.percent <- beds_main_origcap$icu.occupied / beds_main_origcap$icu.cap.orig
``` 

Now remake Heatmap matrices
```{r}
# Inpatient Percent of Capacity
inpat.percent.orig <- spread(beds_main_origcap[,c(1:2,4)],state,inpatient.percent)
inpat.percent.orig.mat <- as.matrix(inpat.percent.orig[,2:50])
rnames <- as.character(inpat.percent.orig[,1])
rownames(inpat.percent.orig.mat) <- rnames

# COVID Percent of Capacity
cov.percent.orig <- spread(beds_main_origcap[,c(1:2,6)],state,covid.percent)
cov.percent.orig.mat <- as.matrix(cov.percent.orig[,2:50])
rnames <- as.character(cov.percent.orig[,1])
rownames(cov.percent.orig.mat) <- rnames

# ICU Percent of Capacity
ICU.percent.orig <- spread(beds_main_origcap[,c(1:2,8)],state,icu.percent)
ICU.percent.orig.mat <- as.matrix(ICU.percent.orig[,2:50])
rnames <- as.character(ICU.percent.orig[,1])
rownames(ICU.percent.orig.mat) <- rnames
```

And remake Heatmaps
```{r}
# Inpatient Percent of Capacity
Heatmap(inpat.percent.orig.mat[,-c(5,19)], # Remove outliers Colorado and Maryland
        row_title = "Date Ascdending from Top", 
        column_title = "Inpatient Percent of Total Beds by State in 2020 \nInitially Reported Inpatient Capacity",
        cluster_rows = FALSE,
        show_row_names = FALSE,
        show_column_dend = FALSE,
        column_names_gp = gpar(fontsize = 10),
        heatmap_legend_param = list(title = "")
        )

# COVID Percent of Capacity
Heatmap(cov.percent.orig.mat[,-c(5,19)], # Remove outliers Colorado and Maryland
        row_title = "Date Ascdending from Top", 
        column_title = "COVID-19 Percent of Total Beds by State in 2020 \nInitially Reported Inpatient Capacity",
        cluster_rows = FALSE,
        show_row_names = FALSE,
        show_column_dend = FALSE,
        column_names_gp = gpar(fontsize = 10),
        heatmap_legend_param = list(title = "")
        )

# ICU Percent of Capacity
Heatmap(ICU.percent.orig.mat[,-c(10,42)], # Remove outliers Georgia and Texas
        row_title = "Date Ascdending from Top", 
        column_title = "Percent of Total ICU Beds by State in 2020 \nInitially Reported ICU Capacity",
        cluster_rows = FALSE,
        show_row_names = FALSE,
        show_column_dend = FALSE,
        column_names_gp = gpar(fontsize = 10),
        heatmap_legend_param = list(title = "")
        )

# Scale drastically increased
# Shows which states had to expand their capacity by the most
```

***

## Results: Regressions

```{r}
icureg <- lm(icu.occupied ~ covid.occupied + inpatient.occupied + collection_date, data = beds_main, na.action = na.omit)
summary(icureg)
  # About 10 additional COVID-19 patients predicts one additional ICU patient
  # About 7 for all inpatient
  # Regression F-stat and p-value indicate that the model has predictive power
  # R^2 is > 90%. 
  # Coefficient p-values indicate high levels of significance

# Test for collinearity
vif(icureg)
```

***

## Discussion/Takeaways

* Inpatient capacity increased drastically at the state- and country-level
  + Far exceeded needs based on increased inpatients at country-level
  + Some states still at or near 100% capacity
* Large increase in inpatients only partially explained by COVID-19 patients
* ICU capacity also increased significantly
  + Occupancy was near capacity pre-COVID-19
  + Increased with COVID, but was more than offset by increased capacity 
* Regression results:
  + Increased COVID-19 inpatients predicts increased ICU occupancy
  + However, to lesser extent than increase in overall inpatients
* States with highest overall inpatient occupancy did not necessarily struggle equally with ICU capacity 

***

## Future Directions

* Continue monitoring capacities to see if they revert to pre-COVID-19 levels
* Investigate increase in non-COVID-19 inpatients. 
  + What were their diagnoses?
  + Was this a reporting issue?
  + If not, how is the increase connected to COVID-19 patients?
* Explore state-level analysis for states most affected by COVID-19
  + Introduce reported COVID-19 cases (not just hospitalizations)
  + Introduce demographic info about states

***

## References

U.S. Department of Health & Human Services, "COVID-19 Estimated Patient Impact and Hospital Capacity by State," HealthData.gov, https://healthdata.gov/dataset/covid-19-estimated-patient-impact-and-hospital-capacity-state

Original data links:

* Inpatient: https://healthdata.gov/sites/default/files/reported_inpatient_all_20200720_0537.csv
* COVID: https://healthdata.gov/sites/default/files/inpatient_covid_final_20200720_0537.csv
* ICU: https://healthdata.gov/sites/default/files/icu_final_20200720_0537.csv

